---
title: "USGS public supply EDA"
author: "Scott Worland"
date: "Tuesday, September 22, 2015"
output: ioslides_presentation
---


## Overview

Two major research objectives:

1. Explore where public supply freshwater withdrawals have changed from 1985-2010 using county level data. 
2. Explore possible environmental, social, and economic drivers of these changes.

Possible journal outlets: 

+ WRR  
+ JAWRA

but...

## Per capita withdrawals

```{r, echo=F, message=F,warning=F,}
library(reshape2);

## load original data
load("USGSWU1985_2010cnty.rda")
d <- WUcnty

## subsample
pop = d[,c(1,2,7,12,17,22,27)]
ground = d[,c(1,3,8,13,18,23,28)]
surface = d[,c(1,4,9,14,19,24,29)]
total = d[,c(1,5,10,15,20,25,30)]

## rename for simpler indexing
names = c("cntyFIPS","1985","1990","1995","2000","2005","2010")
colnames(pop) = names
colnames(ground) = names
colnames(surface) = names
colnames(total) = names

pop2 = melt(pop, variable.name = "year", value.name = "pop")
ground2 = melt(ground, variable.name = "year", value.name = "ground")
surface2 = melt(surface, variable.name = "year", value.name = "surface")
total2 = melt(total, variable.name = "year", value.name = "total")

d2 = cbind(pop2,ground2[,3],surface2[,3],total2[,3])
colnames(d2)[4:6] = c("ground","surface","total")
d3 = melt(d2, id.vars = c("cntyFIPS", "year","pop"),variable.name = "type", value.name="withdrl")
d3$wn = d3$withdrl/d3$pop * 1000
```


```{r dev.args = list(bg = 'transparent'), echo=F, message=F,warning=F, fig.align='center', fig.width=8,fig.height=5.5}
library(ggplot2)

p <- ggplot(d3) 
p <- p + geom_hline(aes(yintercept=3), color="black",linetype = 2, size = 0.5) 
p <- p + geom_hline(aes(yintercept=0), color="black",linetype = 2, size = 0.5)
p <- p + geom_boxplot(aes(type, log10(wn) , fill = year)) + theme_bw(base_size=16)
p
```

## Exploratory data analysis of public supply data

+ Analyze the very high and very low per capita withdrawals
+ Publish EDA findings in OFR/SIR to reference in journal article
+ This presentation contains some of the preliminary investigation

## 2010 National withdrawals
```{r, echo=F, message=F, warning=F,}
theme_blank <- function(){
  require(grid)
  theme(panel.background = element_rect(fill = "transparent",color=NA),
        plot.background = element_rect(fill = "transparent",color=NA),
        legend.background = element_rect(fill = "transparent",color=NA),
        text = element_text(size=20),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.margin=unit(c(0,0,-1,-1),"lines"))
}
```

```{r, echo=F, message=F, warning=F, fig.align='center', cache=T, results='hide'}
library(rgdal); library(ggmap); library(RColorBrewer)

counties <- readOGR("C:\\Users\\scworlan\\Documents\\Water Conservation\\R_conservation\\MSA\\county_shapefiles",
                    layer="cb_2013_us_county_500k");
counties2 <- spTransform(counties, CRS("+proj=longlat + datum=WGS84"));
county.fips <- fortify(counties2, region="GEOID");
colnames(county.fips)[7] <- "GEOID";
counties3 <- merge(county.fips,counties2@data,by="GEOID");
cnty.poly <- subset(counties3, long > -130 & long < 0 & lat > 20 & lat < 50);
colnames(cnty.poly)[1] <- "cntyFIPS";
rm(counties,counties2,county.fips,counties3) # remove old county files for space

state.poly <- map_data("state")

d4 <- subset(d3, year==2010 & type == "total")
d4$brk <- cut(d4$wn, breaks=c(0, 50, 100, 200, 500, 750, 1000, 10000, 25000), 
              labels=c("0-50", "50-100", "100-200", "200-500", "500-750", "750-1k", "1k-10k", ">10k"), 
              include.lowest=TRUE)

d4$brk2 <- cut(d4$withdrl, breaks=c(0, 5, 10, 20, 40, 80, 160, 320, 640), 
              labels=c("0-5", "5-10", "10-20", "20-40", "40-80", "80-160", "160-320", ">320"), 
              include.lowest=TRUE)

cnty.WU <- merge(cnty.poly,d4,by="cntyFIPS")
cnty.WU <- cnty.WU[order(cnty.WU$order),]
```

```{r dev.args = list(bg = 'transparent'), echo=F, message=F, fig.width=9.5, fig.height=6.5, fig.align='center', cache=T}
## Lower 48 raw
m = ggplot(data=cnty.WU) + coord_fixed(1.3)
m = m + geom_polygon(aes(long,lat, group=group,fill=brk2), color=NA)
m = m + geom_polygon(data=state.poly,aes(long,lat, group=group), color="grey", fill=NA, size=0.5)
m = m + scale_fill_manual(values=rev(brewer.pal(8, 'YlGnBu')))
m = m + labs(fill="Mgal/day") + theme_blank()
m
```

## 2010 National per capita withdrawal
```{r dev.args = list(bg = 'transparent'), echo=F, message=F, fig.width=9.5, fig.height=6.5, fig.align='center', cache=T}
## Lower 48
m = ggplot(data=cnty.WU) + coord_fixed(1.3)
m = m + geom_polygon(aes(long,lat, group=group,fill=brk), color=NA)
m = m + geom_polygon(data=state.poly,aes(long,lat, group=group), color="white", fill=NA,size=0.5)
m = m + scale_fill_manual(values=rev(brewer.pal(8, 'YlGnBu')))
m = m + labs(fill="gal/p/d") + theme_blank()
m
```

## 2010 New York withdrawals
```{r, echo=F, message=F, warning=F, fig.align='center', cache=T, results='hide'}
## New york
NY.WU <- subset(cnty.WU, STATEFP==36)
NY.poly <- subset(state.poly, region == "new york")
NJ.poly <- subset(state.poly, region == "new jersey")
PA.poly <- subset(state.poly, region == "pennsylvania")

## NY city watershed (catskills)
NYCws <- readOGR("C:\\Users\\scworlan\\Documents\\Water Conservation\\R_conservation\\MSA\\NY_watershed",
                    layer="New_York_City_Watershed");
NYCws2 <- spTransform(NYCws, CRS("+proj=longlat + datum=WGS84"));
NYCws2@data$id = rownames(NYCws2@data)
NYC.ws <- fortify(NYCws2, region="id");

## Deleware River basin shapefile
DRB <- readOGR("C:\\Users\\scworlan\\Documents\\Water Conservation\\R_conservation\\MSA\\DRB_shapefile",
                 layer="drb_bnd_polygon");
DRB2 <- spTransform(DRB, CRS("+proj=longlat + datum=WGS84"));
DRB2@data$id = rownames(DRB2@data)
DRB.poly <- fortify(DRB2, region="id");
```


```{r dev.args = list(bg = 'transparent'), echo=F, message=F, fig.width=8, fig.height=5, fig.align='center', cache=T}
## withdrawals
m = ggplot(data=NY.WU) + coord_fixed(1.3)
m = m + geom_polygon(aes(long,lat, group=group,fill=brk2), color="grey", size=0.125)
m = m + geom_polygon(data=NY.poly,aes(long,lat, group=group), color="black", fill=NA,size=0.5)
m = m + scale_fill_manual(values=rev(brewer.pal(8, 'YlGnBu')))
m = m + labs(fill="Mgal/day") + theme_blank()
m
```

## 2010 New York per capita withdrawals
```{r dev.args = list(bg = 'transparent'), echo=F, message=F, fig.width=8, fig.height=5, fig.align='center', cache=T}
## normalized withdrawals
m = ggplot(data=NY.WU) + coord_fixed(1.3)
m = m + geom_polygon(aes(long,lat, group=group,fill=brk), color="grey", size=0.125)
m = m + geom_polygon(data=NY.poly,aes(long,lat, group=group), color="black", fill=NA,size=0.5)
m = m + scale_fill_manual(values=rev(brewer.pal(5, 'YlGnBu')))
m = m + labs(fill="gal/p/d") + theme_blank()
m
```

## Big users in New York

```{r, echo=F, message=F}
### clunky way to find high values
high.NY <- data.frame(NY.WU[which(NY.WU$wn>1000),])
high.NY <- high.NY[,c(1,12,17,19,20)]
high.NY <- high.NY[!duplicated(high.NY),]
rownames(high.NY) = NULL
knitr::kable(high.NY, align='c', digits=5,format = "markdown")
```

## From USGS documentation

"There are three major transfers of water out of the Delaware River Basin for New York City.

The NY part of the DRB is mostly in __Delaware__ and __Sullivan__ counties. Other NY counties that also include parts of the DRB are Broome, Greene, __Ulster__, Orange; and also very small parts of Chenango and __Schoharie__ Counties."

## 2010 New York per capita withdrawals

```{r dev.args = list(bg = 'transparent'), echo=F, message=F, fig.width=8, fig.height=5, fig.align='center', cache=T}
## Deleware river basin
m = ggplot(data=NY.WU) + coord_fixed(1.3)
m = m + geom_polygon(aes(long,lat, group=group,fill=brk), color="grey", size=0.125)
m = m + geom_polygon(data=NY.poly,aes(long,lat, group=group), color="black", fill=NA,size=0.5)
m = m + geom_polygon(data=NJ.poly,aes(long,lat, group=group), color="black", fill="grey",size=0.5)
m = m + geom_polygon(data=PA.poly,aes(long,lat, group=group), color="black", fill="grey",size=0.5)
m = m + geom_polygon(data=DRB.poly, aes(long,lat, group=group), fill="darkorchid1",alpha=0.5, color = "darkorchid4")
m = m + scale_fill_manual(values=rev(brewer.pal(5, 'YlGnBu')))
m = m + labs(fill="gal/p/d") + theme_blank()
m
```

## 2010 New York per capita withdrawals

```{r dev.args = list(bg = 'transparent'), echo=F, message=F, fig.width=8, fig.height=5, fig.align='center', cache=T}
## normalized withdrawals
m = ggplot(data=NY.WU) + coord_fixed(1.3)
m = m + geom_polygon(aes(long,lat, group=group,fill=brk), color="grey", size=0.125)
m = m + geom_polygon(data=NYC.ws, aes(long,lat, group=group), fill="darkorchid1",alpha=0.7, color = "darkorchid4")
m = m + geom_polygon(data=NY.poly,aes(long,lat, group=group), color="black", fill=NA,size=0.5)
m = m + scale_fill_manual(values=rev(brewer.pal(5, 'YlGnBu')))
m = m + labs(fill="gal/p/d") + theme_blank()
m
```



## 2010 Texas withdrawals

```{r dev.args = list(bg = 'transparent'), echo=F, message=F, fig.width=8, fig.height=5, fig.align='center', cache=T}
## Texas
TX.WU <- subset(cnty.WU, STATEFP==48)
TX.poly <- subset(state.poly, region == "texas")
TX.cities <- geocode(c("Houston, TX", "San Antonio, TX","Dallas, TX", "Austin, TX"))

## withdrawals
m = ggplot(data=TX.WU) + coord_fixed(1.3)
m = m + geom_polygon(aes(long,lat, group=group,fill=brk2), color="grey", size=0.125)
m = m + geom_polygon(data=TX.poly,aes(long,lat, group=group), color="black", fill=NA,size=0.5)
m = m + geom_point(data=TX.cities, aes(lon, lat), color = "black", fill="red", size = 4, pch=23)
m = m + scale_fill_manual(values=rev(brewer.pal(8, 'YlGnBu')))
m = m + labs(fill="Mgal/day") + theme_blank()
m
```

## 2010 Texas per capita withdrawals

```{r dev.args = list(bg = 'transparent'), echo=F, message=F, fig.width=8, fig.height=5, fig.align='center', cache=T}
## normalized withdrawals
m = ggplot(data=TX.WU) + coord_fixed(1.3)
m = m + geom_polygon(aes(long,lat, group=group,fill=brk), color="grey", size=0.125)
m = m + geom_polygon(data=TX.poly,aes(long,lat, group=group), color="black", fill=NA,size=0.5)
m = m + geom_point(data=TX.cities, aes(lon, lat), color = "black", fill="red", size = 4, pch=23)
m = m + scale_fill_manual(values=rev(brewer.pal(8, 'YlGnBu')))
m = m + labs(fill="gal/p/d") + theme_blank()
m
```

## Big users in Texas
```{r, echo=F, message=F}
### clunky way to find high values
high.TX <- data.frame(TX.WU[which(TX.WU$wn>2000),])
high.TX <- high.TX[,c(1,12,17,19,20)]
high.TX <- high.TX[!duplicated(high.TX),]
rownames(high.TX) = NULL
knitr::kable(high.TX, align='c', digits=5,format = "markdown")
```

